### 2010-08-05 Thesis/Chapter3.tex
- Update 5 August -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@142 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\medskip

@@

	\subsection{Geometry interpolation}
Let us assume a relation between the global coordinate $ x $ and the natural coordinate $ \xi $ in the following form:
\begin{equation}
\label{chap3:f}
x = f(\xi)
\end{equation}
where $ f $ is assumed to be a one-to-one correspondence (that is, a bijective function). This function may be seen as a transformation between the natural shape of the element and its arbitrary shape within the actual mesh, it describes a change in geometry. It is natural to think of approximating the geometry in a similar way that we approximated the solution in section \ref{chap3:solutionInterpolation}. Hence, the transformation defined by \eqref{chap3:f} may also be written as
\begin{equation}
x = \sum_{i=1}^m x_i^e \hat{\psi}_i^e(\xi)
\end{equation}
where $ x_i^e $ is the global coordinate of the $ i^{th} $ node of the element $ \Omega_e $ and $ \hat{\psi}_i^e(\xi) $ are the interpolation functions of degree $ m-1 $. Thus, we have a linear transformation when $ m = 2 $ and the relation betweeen $ x $ and $ \xi $ is quadratic when $ m=3 $. The interpolation functions $ \hat{\psi}_i^e(\xi) $ are called \emph{shape functions} because they are used to express the geometry or shape of the element. 
% Assuming the numbering of nodes in the global system follows the one of the natural system, the shape functions act as a mapping between the two representations.

@@

	\subsection{Time dependence}
In all the equations of the finite element method that we derived so far, we have not taken time into account. For some problem, this is not an issue, the motion is sufficiently slow to assume that the dynamic effects (such as damping effect) are neglectable. However, this assumption is not always valid, particulary in our case where we want to model organ deformation. The deformation obviously depends on time. Finite element models of time-dependent problems can be developed in two alternatives ways \citep{Reddy93}:
\begin{description}
\item[(a)] coupled formulation in which the time $ t $ is treated as an additional coordinate along with the spatial coordinate $ x $
\item[(b)] decoupled formulation where time and spatial variations are assumede to be separable.
\end{description}
Thus, the approximation of the solution $ u $ takes one of these two forms:
\begin{align}
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n \hat{u}^e_i \hat{\psi}^e_i(x, t)  \quad \text{(coupled formulation)} \label{chap3:approxTime1}\\
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n u^e_i(t) \psi^e_i(x) \quad \text{(decoupled formulation)} \label{chap3:approxTime2}
\end{align}
where $ \hat{\psi}^e_i(x, t) $ are time-space interpolation functions, $ \hat{u}^e_i $ are the nodal values that are independent of $ x $ and $ t $, $ \psi^e_i(x) $ are the usual interpolation functions in spatial coordinate $ x $ only and the nodal values $ u^e_i(t) $ are functions of time $ t $ only. Space-time coupled finite element formulations are not common and they have not been adequately studied. Hence, we consider the decoupled formulation only. Of course, the assumption that the time and spatial variations are separable is not valid in general. However, with sufficiently small time steps, it is possible to obtain accurate solutions nevertheless.

@@

	\subsection{Determination of the matrices involved}

%	\subsection{Generation of element matrices}
%The basic postulate is that a continuum must satisfy Newton's second law of motion. This law can be stated as: the time rate of change of momentum of a collection of particules equals the resultant force exerted on the collection:
%\begin{equation}
%\dfrac{d}{dt}(m \mathbf{v}) = \mathbf{F},
%\end{equation}
%where $ m $ is the total mass, $ \mathbf{v} $ the velocity and $ \mathbf{F} $ the resultant force on the collection of particules.

\section{Assembly of element equations}
%Since the element is physically connected to its neighbours, the resulting algebraic equations will contain more unknowns than the number of algebraic equations. Thus, it becomes necessary to put the elements together to eliminate the extra unknowns. This process is described in the next section.

>>>>

\bigskip

@@

	\subsection{Geometry interpolation}
Let us assume a relation between the global coordinate $ x $ and the natural coordinate $ \xi $ in the following form:
\begin{equation}
\label{chap3:f}
x = f(\xi)
\end{equation}
where $ f $ is assumed to be a one-to-one correspondence (that is, a bijective function). This function may be seen as a transformation between the natural shape of the element and its arbitrary shape within the actual mesh, it describes a change in geometry. It is natural to think of approximating the geometry in a similar way that we approximated the solution in section \ref{chap3:solutionInterpolation}. Hence, the transformation defined by \eqref{chap3:f} may also be written as
\begin{equation}
\label{chap3:shapeFunctions}
x = \sum_{i=1}^m x_i^e \hat{\psi}_i^e(\xi)
\end{equation}
where $ x_i^e $ is the global coordinate of the $ i^{th} $ node of the element $ \Omega_e $, $ m $ the number of nodes for the element and $ \hat{\psi}_i^e(\xi) $ are the interpolation functions of degree $ m-1 $. Thus, we have a linear transformation when $ m = 2 $ and the relation betweeen $ x $ and $ \xi $ is quadratic when $ m=3 $. The interpolation functions $ \hat{\psi}_i^e(\xi) $ are called \emph{shape functions} because they are used to express the geometry or shape of the element. 
% Assuming the numbering of nodes in the global system follows the one of the natural system, the shape functions act as a mapping between the two representations.

\bigskip

\noindent \textbf{Remark.}
It is worth noting that \eqref{chap3:shapeFunctions} can be easily extented to three-dimensional problems. Let us consider a Cartesian coordinate system and we denote the position of node $ i $ $ \mathbf{x}_i = [x_{i1}, x_{i2}, x_{i3}]^T$. The position of a point $ \mathbf{x} = [x_{1}, x_{2}, x_{3}]^T$ within the element may then be interpolated by applying \eqref{chap3:shapeFunctions} to each component of the position:
\begin{align}
x_1 &= \sum_{i=1}^m x_{i1} \hat{\psi}_i^e(\xi) \notag \\
x_2 &= \sum_{i=1}^m x_{i2} \hat{\psi}_i^e(\xi) \\
x_3 &= \sum_{i=1}^m x_{i3} \hat{\psi}_i^e(\xi) \notag
\end{align}
If we construct the vector $ \hat{\mathbf{x}}^e $ of all nodal positions $ \mathbf{x}_i $ of the element, we can define a shape function matrix $ \mathbf{H} $ such as:
\begin{equation}
\label{chap3:H}
\mathbf{x} = \mathbf{H} \hat{\mathbf{x}}^e
\end{equation}
where $ \mathbf{H} $ is a $ 3 \times m $ matrix built from submatrices $ \mathbf{H}_i $
\begin{equation}
\mathbf{H}_i = 
	\begin{bmatrix}
		\hat{\psi}_i^e & 0 & 0 \\
		0 & \hat{\psi}_i^e & 0 \\
		0 & 0 & \hat{\psi}_i^e
	\end{bmatrix}
\quad i = 1, 2, \ldots, m.
\end{equation}

@@

	\subsection{Time dependence}
In all the equations of the finite element method that we derived so far, we have not taken time into account. For some problem, this is not an issue, the motion is sufficiently slow to assume that the dynamic effects (such as damping effect) are neglectable. However, this assumption is not always valid, particulary in our case where we want to model organ deformation. The deformation obviously depends on time. Finite element models of time-dependent problems can be developed in two alternatives ways \citep{Reddy93}:
\begin{description}
\item[(a)] coupled formulation in which the time $ t $ is treated as an additional coordinate along with the spatial coordinate $ x $
\item[(b)] decoupled formulation where time and spatial variations are assumede to be separable.
\end{description}
Thus, the approximation of the solution $ u $ takes one of these two forms:
\begin{align}
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n \hat{u}^e_i \hat{\psi}^e_i(x, t)  \quad \text{ (coupled formulation)} \label{chap3:approxTime1}\\
u(x, t) \approx u^e(x, t) &= \sum_{i=1}^n u^e_i(t) \psi^e_i(x) \quad \text{(decoupled formulation)} \label{chap3:approxTime2}
\end{align}
where $ \hat{\psi}^e_i(x, t) $ are time-space interpolation functions, $ \hat{u}^e_i $ are the nodal values that are independent of $ x $ and $ t $, $ \psi^e_i(x) $ are the usual interpolation functions in spatial coordinate $ x $ only and the nodal values $ u^e_i(t) $ are functions of time $ t $ only. Space-time coupled finite element formulations are not common and they have not been adequately studied. Hence, we consider the decoupled formulation only. Of course, the assumption that the time and spatial variations are separable is not valid in general. However, with sufficiently small time steps, it is possible to obtain accurate solutions nevertheless.

@@

	\subsection{A few words on the matrices involved}

		\subsubsection{Mass matrix}
The mass matrix $ \mathbf{M} $ for the whole domain may be computed by summing the mass contributions from every elements:
\begin{equation}
\mathbf{M} = \sum_e \mathbf{M}^e,
\end{equation}
where $ \mathbf{M}^e $ is the mass of element $ \Omega_e $ given by:
\begin{equation}
\mathbf{M}^e = \int_{V^e} \rho^e \mathbf{H}^T \mathbf{H} dV
\end{equation}
where $ \rho $ the mass density, $ V^e $ the volume of the element and $ \mathbf{H} $ is the shape function matrix defined by \eqref{chap3:H}. This integral may be evaluated using a procedure of numerical integration and generally results in a band mass matrix (non-zero entries are confined to a diagonal band, comprising the main diagonal and more diagonals on either side). This method for computing the mass matrix is called \emph{variational mass lumping}.

However, it is very common in finite element analysis to use a diagonal mass matrix. The most common way of achieving this is to define the diagonal entries with the sum of the corresponding rows and then set all other values to 0. The physical interpretation of this is lumping all of the mass in the volume surrounding a node at the node itself. A diagonal mass matrix may offer significant computational and storage advantages. First, it is easily inverted, since the inverse of a diagonal matrix is also diagonal. And a diagonal mass matrix can also be stored simply as a vector. Therefore, such a matrix may be built by directly computing the mass of each element and assigning an equal proportion of this to each of the element's nodes. This method is named \emph{direct mass lumping}. The mass $ m^e $ of element $ \Omega_e $ is given by:
\begin{equation}
m^e = \rho^e \int_{V^e} dV,
\end{equation}
which is integrated over the element using a numerical method. Each diagonal component of the mass matrix then receives a contribution of $ m^e/m $.

		\subsubsection{Damping matrix}
In a similar way to the mass matrix, a damping matrix may be defined by summing the damping contributions from every elements:
\begin{equation}
\mathbf{D} = \sum_e \mathbf{D}^e,
\end{equation}
where $ \mathbf{D}^e $ is the damping of element $ \Omega_e $ given by:
\begin{equation}
\mathbf{D}^e = \int_{V^e} \kappa^e \mathbf{H}^T \mathbf{H} dV
\end{equation}
where $ \kappa^e $ is a damping parameter. In practice, the components of damping matrices are difficult to determine experimentally. Therefore, it is common to employ the \emph{Rayleigh damping}. The Rayleigh damping assumes that the damping may be written in the form of:
\begin{equation}
\mathbf{D}^e = \alpha \mathbf{M}^e + \beta \mathbf{K}^e,
\end{equation}
where $ \alpha $ and $ \beta $ are coefficients defining mass and stiffness proportional damping components. However, for systems with large degrees of freedom, it is difficult to guess meaningful values of $ \alpha $ and $ \beta $. \cite{Chowdhury03} have proposed a method to calculate these coefficients. In practice, further assumptions may only lead to include the mass proportional component ($ \beta = 0 $), which yields a diagonal damping matrix if a diagonal mass matrix is used.

		\subsubsection{Stiffness matrix}
The expression of the stiffness matrix depends on the chosen strain tensor as well as the appropriate constitutive model for the material. Its general form cannot be explicited. However, it can be demonstrated that $ \mathbf{K}^e $ may be written as
\begin{equation}
\mathbf{K}^e = \int_{V^e} \mathbf{B}^T \mathbf{M} \, \mathbf{B} \, dV,
\end{equation}
where $ \mathbf{M} $ is a material matrix, which only depends on the constitutive model chosen, and $ \mathbf{B} $ is the \emph{strain-displacement matrix} defined as the derivatives of the shape functions (see \cite{Liu03,Belytschko00} for details).

\section{Assembly of element equations}
We have now derived the equations for each element. However, since the element is physically connected to its neighbours, the resulting algebraic equations will contain more unknowns than the number of algebraic equations. Thus, it becomes necessary to put the elements together to eliminate the extra unknowns. The assembly should be carried out as soon as the element matrices are computed, rather than waiting until the element coefficients of all elements are computed. This would require storage of the coefficients for each element. Instead, we can perform the assembly while we calculate the element matrices.

The assembly operation is actually quite simple. Let us consider an example with stiffness matrices for one-dimensional problem. We assume that we have three two-node elements. And we suppose the element connectivity information given by Table \ref{chap3:connectivity}, which provides the global node number for each local node number within an element. 
\begin{table}
	\begin{center}
	\begin{tabular}{|c|c|c|}
		\hline Elements & 1 & 2 \\ 
		\hline 1 & 1 & 2 \\ 
		\hline 2 & 2 & 3 \\ 
		\hline 3 & 3 & 4 \\ 
		\hline 
		\end{tabular} \caption{Element connectivity information}
		\label{chap3:connectivity}
	\end{center}
\end{table}

% The element matrices are expressed with respect to local node numbers.
\TODO{add figure with global stiffness matrix and a few arrows issued from the element stiffness matrices}
